Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Worked on background.

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}[1]{\mathsf{get\_group\_id(#1)}}
\newcommand{\getnumgroups}[1]{\mathsf{get\_num\_groups(#1)}}

@@

\ADComment{Use the work stealing example to both introduce OpenCL and
  also make our key point: that arbitrary preemption will not work.}

@@

Our first additional language primitive, $\offerkill$, is a
workgroup-level function: similar to the OpenCL intra-workgroup
$\mathsf{barrier}$ primitive, if $\offerkill$ appears inside
conditional code then it must be executed by \emph{all} threads in a
workgroup, with all threads having executed the same number of
iterations of any enclosing loops.

>>>>

\newcommand{\offerfork}{\mathsf{offer\_fork}}
\newcommand{\offerkill}{\mathsf{offer\_kill}}
\newcommand{\globalbarrier}{\mathsf{global\_barrier}}
\newcommand{\resizingglobalbarrier}{\mathsf{resizing\_global\_barrier}}
\newcommand{\getgroupid}{\mathsf{get\_group\_id}}
\newcommand{\getnumgroups}{\mathsf{get\_num\_groups}}
\newcommand{\getlocalid}{\mathsf{get\_local\_id}}
\newcommand{\getglobalid}{\mathsf{get\_global\_id}}
\newcommand{\getlocalsize}{\mathsf{get\_global\_id}}

@@

We use two worked examples---work stealing and breadth-first graph
search---to provide background on the OpenCL programming model and
irregular data-parallel algorithms.

\begin{figure}
\ADComment{Tyler/Hugues: please insert workstealing example.}
\begin{lstlisting}
TODO
\end{lstlisting}
\caption{An excerpt of a work stealing algorithm in OpenCL}\label{fig:workstealing}
\end{figure}

\begin{figure}
\ADComment{Tyler/Hugues: please insert graph traversal example.}
\begin{lstlisting}
TODO
\end{lstlisting}
\caption{An OpenCL graph search algorithm, using a global barrier for synchronisation}\label{fig:graphsearch}
\end{figure}

\paragraph{OpenCL}
An OpenCL program is divided into \emph{host} and \emph{device}
components.  A host application runs on the CPU and launches one or
more \emph{kernels} that run on accelerator devices---GPUs in the
context of this paper.  A kernel is writen in OpenCL C, based on C99.
All threads executing a kernel execute the same entry function with
identical arguments, but a thread may query its unique thread id in
order to access different data or follow different control flow paths
to other threads.

Figure~\ref{fig:workstealing} shows a simplified version of an OpenCL
kernel that uses a work stealing queue.  \ADComment{Describe some
  aspects of the example, saying that the mutex is not a primitive,
  and can be coded up using the memory model but is not safe due to
  the execution model.  Make it clear that the lock function uses busy-waiting.}  Figure~\ref{fig:graphsearch} shows the
skeleton of a graph traversal algorithm.  \ADComment{Describe some
  aspects of the example, saying that the global barrier is not a
  primitive, and can be coded up using the memory model but is not
  safe due to the execution model.  Make it clear that the barrier uses busy-waiting.}

\paragraph{Execution model}

The threads executing a kernel are hierarchically organised.  A kernel
is executed by a number of \emph{workgroups}, and each workgroup
consists of a fixed number of \emph{threads}.  Within a workgroup,
threads may also be organised into \emph{subgroups} (termed
\emph{warps} and \emph{wavefronts} on NVIDIA and AMD architectures,
respectively).

Two barrier primitives are provided in OpenCL.  A workgroup barrier
synchronises all threads in a workgroup and ensures memory
consistency.  This barrier is a \emph{workgroup-level} function: it
must be encountered uniformly by all or none of the threads of a
workgroup, thus can only appear in conditional code if all threads in
a workgroup that reach the barrier agree on the guards of all
enclosing conditional statements.  A subgroup barrier works
analogously at the subgroup level.  As discussed below, a
\emph{global} barrier (synchronising all threads running a kernel) is
not provided as a primitive.

OpenCL specifically makes no guarantees about fair scheduling between
workgroups, stating~\cite{...}: \emph{``A conforming implementation
  may choose to serialize the workgroups so a correct algorithm cannot
  assume that workgroups will execute in parallel.  There is no safe
  and portable way to synchronize across the independent execution of
  workgroups since once in the work-pool, they can execute in any
  order.''}  OpenCL is not alone in this regard: CUDA similarly
provides no guarantees~\cite{...}, and HSA guarantees only that
\ADComment{Tyler, can you state the HSA guarantee?}~\cite{...}.

Revisiting the examples of Figures~\ref{fig:workstealing}
and~\ref{fig:graphsearch}: the mutex function of
Figure~\ref{fig:workstealing} will cause starvation if a workgroup
that holds the mutex is not scheduled again and so does not release
the mutex.  A similar issue affects the global barrier function of
Figure~\ref{fig:graphsearch}.  Because the barrier is implemented
using busy-waiting, a set of workgroups waiting at the barrier may be
continuously scheduled, starving a workgroup that has not yet arrived
at the barrier and meaning that no workgroup proceeds past the
barrier.

\paragraph{Id functions}

Builtin functions $\getlocalid()$ and $\getglobalid()$ provide a
thread with an id that is unique within its workgroup and unique
across all workgroups, respectively.  Each workgroup has an id,
obtained via $\getgroupid()$.  The number of threads per workgroup and
the number of workgroups are obtained via $\getlocalsize()$ and
$\getnumgroups()$.  A thread's local and global id are related via the
equation $\getglobalid() = \getgroupid() \times \getlocalsize() +
\getlocalid()$.

In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above id functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.

\paragraph{Memory model}

An OpenCL kernel has access to four memory spaces.  Shared virtual
memory (SVM) is accessible to all threads, and to the host
application.  Atomic operations on this memory space enable
fine-grained communication between host and device.  Our
implementation of cooperative kernels
(Section~\ref{sec:implementation}) depends on this recent OpenCL
feature.  Global memory is shared among all threads executing a
kernel, but can only be safely accessed by the host application before
or after kernel execution.  Local memory is shared among all threads
in the same workgroup: each workgroup has its own portion of local
memory, which can be used for fast intra-workgroup communication.
Each thread has an allocation of very fast private memory, where
function-local variables are allocated.

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.
\ADComment{Perhaps say more.}  \ADComment{Can we illustrate using the
  examples?}

@@

Similar to the OpemncL workgroup $\mathsf{barrier}$ primitive, our first additional language primitive, $\offerkill$, is a
workgroup-level function (see Section~\ref{sec:background}).
